### 2024-11-07 Analysis/01_Simulating_spaces_and_stressors.Rmd
added precision on mechanism expectations

```{r, eval = FALSE}
plot.text.box <- function(my_text, ...) {
  plot(1,1, xaxt = "n", yaxt = "n", xlab = "", ylab = "", col = "white", pch = 21)
  text(c(1,1), my_text, ...)
}

@@

## Stressors
plot.text.box("B) Resampling\nwith stressor:\nnull mechanism", cex = 2)
plot.1D(trait_space, reduction = random_reduction, null = !random_reduction, main = "1D null mechanism")
plot.2D(trait_space, reduction = random_reduction, null = !random_reduction, main = "2D null mechanism")

plot.text.box("C) Resampling\nwith stressor:\nequalizing\nmechanism", cex = 2)
plot.1D(trait_space, reduction = equalizing_fitness, null = random_reduction, main = "1D equalising mechanism")
plot.2D(trait_space, reduction = equalizing_fitness, null = random_reduction, main = "2D equalising mechanism")

plot.text.box("D) Resampling\nwith stressor:\nfacilitating\nmechanism", cex = 2)
plot.1D(trait_space, reduction = facilitation, null = random_reduction, main = "1D facilitation mechanism")
plot.2D(trait_space, reduction = facilitation, null = random_reduction, main = "2D facilitation mechanism")

plot.text.box("E) Resampling\nwith stressor:\nfiltering\nmechanism", cex = 2)
plot.1D(trait_space, reduction = filtering_reduction, null = random_reduction, main = "1D filtering mechanism")
plot.2D(trait_space, reduction = filtering_reduction, null = random_reduction, main = "2D filtering mechanism")

plot.text.box("F) Resampling\nwith stressor:\ncompetition\nmechanism", cex = 2)
plot.1D(trait_space, reduction = !competitive_exclusion, null = random_reduction, main = "1D competition mechanism")
plot.2D(trait_space, reduction = !competitive_exclusion, null = random_reduction, main = "2D competition mechanism")

>>>>

```{r, eval = TRUE}
plot.text.box <- function(my_text, ...) {
  plot(1,1, xaxt = "n", yaxt = "n", xlab = "", ylab = "", col = "white", pch = 21)
  text(c(1,1), my_text, ...)
}

@@

## Stressors
plot.text.box("B) Null mechanism:\nno specific\nmetric change", cex = 2)
plot.1D(trait_space, reduction = random_reduction, null = !random_reduction, main = "1D null mechanism")
plot.2D(trait_space, reduction = random_reduction, null = !random_reduction, main = "2D null mechanism")

plot.text.box("C) Approximating\nequalizing\nmechanism:\nexpected changes\nin richness", cex = 2)
plot.1D(trait_space, reduction = equalizing_fitness, null = random_reduction, main = "1D equalising mechanism")
plot.2D(trait_space, reduction = equalizing_fitness, null = random_reduction, main = "2D equalising mechanism")

plot.text.box("D) Approximating\nfacilitating\nmechanism:\nexpected changes\nin regularity", cex = 2)
plot.1D(trait_space, reduction = facilitation, null = random_reduction, main = "1D facilitation mechanism")
plot.2D(trait_space, reduction = facilitation, null = random_reduction, main = "2D facilitation mechanism")

plot.text.box("E) Approximating\nfiltering\nmechanism:\nexpected changes\nin position", cex = 2)
plot.1D(trait_space, reduction = filtering_reduction, null = random_reduction, main = "1D filtering mechanism")
plot.2D(trait_space, reduction = filtering_reduction, null = random_reduction, main = "2D filtering mechanism")

plot.text.box("F) Approximating\ncompetition\nmechanism:\nexpected changes\nin divergence", cex = 2)
plot.1D(trait_space, reduction = !competitive_exclusion, null = random_reduction, main = "1D competition mechanism")
plot.2D(trait_space, reduction = !competitive_exclusion, null = random_reduction, main = "2D competition mechanism")

### 2024-11-07 Manuscript/main_ecography_R1.tex
added precision on mechanism expectations

\section*{Box 1: what, how and why = pattern, process and mechanism}
\cite{anand1994pattern} conceptualised a general analytical framework for answering scientific questions in ecology and evolution, based on three sequential questions: \textit{``what?''} (describing the pattern), \textit{``how?''} (describing the process) and \textit{``why?''} (understanding the mechanism).
This framework can be effectively applied to trait-based analyses:

@@

\begin{itemize}
\item Random removal: by randomly removing 20\%, 40\%, 60\% or 80\% of the data.
This approximates our \textbf{null mechanism}. This is used to establish a reference to be compared to the other mechanisms and to test how removing elements in a specific way influences the metrics scores compared to removing them randomly.

\item Decreasing size: by removing the required amount (20\%, 40\%, 60\% or 80\%) of data away from a distance (radius) $\rho$ of the centre of the trait space.
$\rho$ is estimated for each trait space to match the required intensity. For example, for one trait space, the algorithm estimated $\rho_{20}$ to be the minimum radius excluding 20\% of the observations, $\rho_{40}$ the one removing 40\%, etc. (with $\rho_{20} > \rho_{40} > \rho_{60} > \rho_{80}$).
This approximates our \textbf{equalising mechanism}.

\item Increasing density: by removing the required amount of pairs of points with a variable pairwise distance of at least $D$.
$D$ is a distance estimated for each trait space to correspond to the minimal Euclidean distance that encompasses at least $n$ pairs of points corresponding to the intensity of the stressor.
For example $D_{20}$ is the distance that excludes any $n_{20}$ pairs of points that are a distance of at least $D_{20}$ from each other, $D_{40}$ is the distance that excludes $n_{40}$ pairs, etc. (with $D_{20} > D_{40} > D_{60} > D_{80}$).
Note that this algorithm is not directly based on the change of average density but rather on the change in pairwise distance between observations.
This approximates our \textbf{facilitation mechanism} (i.e. the points left are only ones that are close to at least another point in space).

\item Shifting space: by removing the required amount of data from a distance (radius) $\rho'$ of the observation with the maximum numerical value on all dimensions.
This is similar to the decreasing size algorithm but instead of choosing $\rho$ to be the radius from the centre of the trait space, $\rho'$ is a radius from the ``top right corner'' of the trait space.
In other words the centre of the radius $\rho'$ is the observation with the highest numerical value on all dimensions - in a 2D representation this is the observation in the top right corner (with $\rho'_{20} > \rho'_{40} > \rho'_{60} > \rho'_{80}$).
This approximates our \textbf{filtering mechanism}.

\item Increasing evenness: by resampling the proportion of data (i.e. 20\%, 40\%, 60\% or 80\%) but with skewed resampling probabilities.
In brief, this algorithm reduces the probability of resampling observations in regions of the trait space that have many observations and increases it in regions that have few observations.
This is done by selecting of $B$ discrete categories to summarise the distribution - i.e. bandwidths - using Silverman's ``rule of thumb'' \citep[\texttt{bw.nrd0()} function in \texttt{R};][]{silverman1986density}.
Each of these discrete category $B$ has an observed probability of sampling of $b$ (the proportion of observations in the category) and each observation then gets a probability of resampling scaling with the intensity of the stressor (the proportion of what to remove $i$) of $i \times (1-b)^{p}$. 
Where $p$ is factor increasing the scaling power of the algorithm (here we used $p=3$).
In other words, observations in categories with a low density ($b<0.5$) were more likely to be resampled and the ones in categories with high density ($b>0.5$) were less likely to be resampled.
This approximates our \textbf{competition mechanism}, were observations in dense regions of the trait space are more likely to be removed than in sparse regions of the trait space.
\end{itemize}

@@

Note that in empirical data, depending on the distribution of the data, some mechanisms can lead to similar or dissimilar patterns.
For example, if the data are normally distributed, the equalising mechanisms, by removing data on the edges of the distribution also increases the density of the trait space (because normally distributed data are denser in the centre of the distribution), similarly to the facilitation mechanism.
However, if the data are distributed uniformly, this does not happen.
Furthermore, in real-world scenarios, we do not expect these mechanisms to act in isolation of each other: multiple mechanisms may stress the observed data simultaneously, with cumulative or synergistic effects.
However, this is not tackled here for both simplicity and to understand how they work in isolation.

@@

\begin{table}
\center
\scriptsize
\begin{tabular}{p{0.2\linewidth}|p{0.2\linewidth}|p{0.4\linewidth}|p{0.2\linewidth}}
\textbf{Approximated mechanism (biological)} & \textbf{Used mechanism (algorithm in \texttt{dispRity::reduce.space})} & \textbf{Stressor description} & \textbf{Type of metric expected to recover the mechanism}\\
\hline
  Null mechanism & Random removal (``random'') & No overall change in observations' traits or community structure except for the reduction in the number of observations. & Metrics sensitive to the amount of data \\
  Equalising fitness & Size change (``size'') & When observations with trait combinations that are more extreme (i.e. observations located away from the centre of the trait space) are disadvantaged due to some resource concentration gradient \citep{chesson2000mechanisms,barot2004mechanisms}. & Metrics capturing changes in richness (but not divergence and regularity) \\
  Facilitation & Density change (``density'') & When observations with closely shared trait combinations (i.e. located close together in the trait space) are more likely to survive the stressor \citep{bruno2003inclusion, danet2024species}. In other words, this simulates the idea that some observations are more likely to be resilient if they are present in a community of observations (e.g. a community of species) with shared traits rather than the opposite. & Metrics capturing changes in regularity (but not richness and regularity) \\
  Environmental filtering & Position change (``position'') & When an increase in observations' trait similarity happens through some strong abiotic constraints \citep{cornwell2006trait}. For example, some trait combinations become more and more unlikely due to some environmental constraints (i.e. some regions of the trait space become unsuitable). & Metrics capturing changes in position (and richness) \\\\
  Competitive exclusion & Evenness change (``evenness'') & When functionally similar observations compete more intensively with one another than with functionally dissimilar observations (``competitive exclusion principle''; \citealt{hardin1960competitive}). At the extreme, it implies that only dissimilar individuals will coexist (``limiting similarity principle''; \citealt{macarthur1967limiting}). & Metrics capturing changes in divergence (but not richness and regularity) \\\\

@@

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.5\textwidth]{Figures/simulation_protocol_explained.pdf}
\caption{\scriptsize{Illustration of the simulation protocol used in this paper.
A) we simulated a pure birth-tree with a trait evolving under a Brownian Motion process until reaching 200 observations resulting a trait space of 2, 4, or 8 dimensions.
The first plot on row A illustrates one such simulation in one dimensions through time (with time on the horizontal axis and trait value for one dimension on the vertical axis).
The second plot on the same row shows the same results but with 2 dimensions of the trait represented here.
We then applied different stressors to the resulting trait space.
We represent the observations removed by the stressors in blue and the ones kept in the trait space in orange.
The first column (histograms) illustrates the distribution of observations in one dimension of the 2D, 4D or 8D traits with the overlap between the removed (blue) and kept (orange) observations in brown (note we did not measure any metric on 1D spaces).
The second column illustrate the trait in 2 dimensions.
We did not illustrate the data beyond 2 dimensions for traits in 4 or 8 dimensions.
We used the following stressors:
B) \textbf{null mechanism}: randomly removing a proportion of the observations;
C) \textbf{approximating equalising mechanism}: removing observations on the edge of the distribution;
D) \textbf{approximating facilitation mechanism}: removing observations to reduce the distance between pairs of observations (increasing local density; decreasing evenness);
E) \textbf{approximating filtering mechanism}: removing observations on one extreme of the distribution;
F) \textbf{approximating competition mechanism}: proportionally removing observations on the centre of the distribution.
G) we presented the results of the simualtions as the scaled metric scores for the approximated mechanisms (C to F) relative to the null mechanism. The dot represents the median metric score and the thick and dashed lines respectively the 50\% and 95\% confidence intervals.}}
\label{Fig:simulations}
\end{figure}
\bigskip

>>>>

\section*{Box 1: what, how and why: pattern, process and mechanism}
\cite{anand1994pattern} conceptualised a general analytical framework for answering scientific questions in ecology and evolution, based on three sequential questions: \textit{``what?''} (describing the pattern), \textit{``how?''} (describing the process) and \textit{``why?''} (understanding the mechanism).
This framework can be effectively applied to trait-based analyses:

@@

\begin{itemize}
\item Random removal: by randomly removing 20\%, 40\%, 60\% or 80\% of the data.
This approximates our \textbf{null mechanism}. This is used to establish a reference to be compared to the other mechanisms and to test how removing elements in a specific way influences the metrics scores compared to removing them randomly. We chose this algorithm to approximate the absence of any specific stressor (i.e. the removal of data only changes the number of observations, not the properties of the trait space).

\item Decreasing size: by removing the required amount (20\%, 40\%, 60\% or 80\%) of data away from a distance (radius) $\rho$ of the centre of the trait space.
$\rho$ is estimated for each trait space to match the required intensity. For example, for one trait space, the algorithm estimated $\rho_{20}$ to be the minimum radius excluding 20\% of the observations, $\rho_{40}$ the one removing 40\%, etc. (with $\rho_{20} > \rho_{40} > \rho_{60} > \rho_{80}$).
This approximates our \textbf{equalising mechanism}. 
We chose this algorithm to approximate equalisation assuming that a stressor could increase the probability of extinction for observations with more rare trait combinations (i.e. observations on the edges of the trait space).

\item Increasing density: by removing the required amount of pairs of points with a variable pairwise distance of at least $D$.
$D$ is a distance estimated for each trait space to correspond to the minimal Euclidean distance that encompasses at least $n$ pairs of points corresponding to the intensity of the stressor.
For example $D_{20}$ is the distance that excludes any $n_{20}$ pairs of points that are a distance of at least $D_{20}$ from each other, $D_{40}$ is the distance that excludes $n_{40}$ pairs, etc. (with $D_{20} > D_{40} > D_{60} > D_{80}$).
Note that this algorithm is not directly based on the change of average density but rather on the change in pairwise distance between observations.
This approximates our \textbf{facilitation mechanism} (i.e. the points left are only ones that are close to at least another point in space).
We chose this algorithm to approximate facilitation mechanism where a stressor could increase the probability of extinction for observations that are more isolated in the trait space.
I.e. pairs of observations that share a close trait combinations are more likely to remain (i.e. observations adjacent to each other in the trait space).

\item Shifting space: by removing the required amount of data from a distance (radius) $\rho'$ of the observation with the maximum numerical value on all dimensions.
This is similar to the decreasing size algorithm but instead of choosing $\rho$ to be the radius from the centre of the trait space, $\rho'$ is a radius from the ``top right corner'' of the trait space.
In other words the centre of the radius $\rho'$ is the observation with the highest numerical value on all dimensions - in a 2D representation this is the observation in the top right corner (with $\rho'_{20} > \rho'_{40} > \rho'_{60} > \rho'_{80}$).
This approximates our \textbf{filtering mechanism}.
We chose this algorithm to approximate a flitering assuming that a stressor that could increase the probability of extinction for observations further from some trait combination optimum.
This is similar to our equalising mechanism approximation but with the optimum not being in the center of the trait space (a region with a high density of observations) but with the optimum being in a corner of the trait space (a region with low density).

\item Increasing evenness: by resampling the proportion of data (i.e. 20\%, 40\%, 60\% or 80\%) but with skewed resampling probabilities.
In brief, this algorithm reduces the probability of resampling observations in regions of the trait space that have many observations and increases it in regions that have few observations.
This is done by selecting of $B$ discrete categories to summarise the distribution - i.e. bandwidths - using Silverman's ``rule of thumb'' \citep[\texttt{bw.nrd0()} function in \texttt{R};][]{silverman1986density}.
Each of these discrete category $B$ has an observed probability of sampling of $b$ (the proportion of observations in the category) and each observation then gets a probability of resampling scaling with the intensity of the stressor (the proportion of what to remove $i$) of $i \times (1-b)^{p}$. 
Where $p$ is factor increasing the scaling power of the algorithm (here we used $p=3$).
In other words, observations in categories with a low density ($b<0.5$) were more likely to be resampled and the ones in categories with high density ($b>0.5$) were less likely to be resampled.
This approximates our \textbf{competition mechanism}, were observations in dense regions of the trait space are more likely to be removed than in sparse regions of the trait space.
We chose this algorithm to approximate a stressor that could increase the probability of extinction for observations that share traits combination.
This could approximate the fact that observation that have similar trait combinations are more likely to go extinct, due to competition (e.g. using the same resources). 
\end{itemize}

@@

Note that in empirical data, depending on the distribution of the data, some mechanisms can lead to similar or dissimilar patterns.
For example, if the data are normally distributed, the equalising mechanisms, by removing data on the edges of the distribution also increases the density of the trait space (because normally distributed data are denser in the centre of the distribution), similarly to the facilitation mechanism.
However, if the data are distributed uniformly, this does not happen.
Furthermore, in real-world scenarios, we do not expect these mechanisms to act in isolation of each other: multiple mechanisms may stress the observed data simultaneously, with cumulative or synergistic effects.
However, this is not tackled here for both simplicity and to understand how they work in isolation. 
Therefore the prediction of how each different mechanism approximation would affect the metrics (Table \ref{Tab:mechanisms}) should be taken as a cautious guideline when dealing with empirical data where the distribution of the data might not be normal and where it could be affected by multiple mechanisms.

@@

\begin{table}
\center
\scriptsize
\begin{tabular}{p{0.2\linewidth}|p{0.2\linewidth}|p{0.4\linewidth}|p{0.2\linewidth}}
\textbf{Approximated mechanism (biological)} & \textbf{Used mechanism (algorithm in \texttt{dispRity::reduce.space})} & \textbf{Stressor description} & \textbf{Type of metric expected to recover the mechanism}\\
\hline
  Null mechanism & Random removal (``random'') & No overall change in observations' traits or community structure except for the reduction in the number of observations. & Metrics sensitive to the amount of data \\

  Equalising fitness & Size change (``size'') & When observations with trait combinations that are more extreme (i.e. observations located away from the centre of the trait space) are disadvantaged due to some resource concentration gradient \citep{chesson2000mechanisms,barot2004mechanisms}. & Metrics capturing changes in richness (but not divergence and regularity) \\

  Facilitation & Density change (``density'') & When observations with closely shared trait combinations (i.e. located close together in the trait space) are more likely to survive the stressor \citep{bruno2003inclusion, danet2024species}. In other words, this simulates the idea that some observations are more likely to be resilient if they are present in a community of observations (e.g. a community of species) with shared traits rather than the opposite. & Metrics capturing changes in regularity (but not richness and regularity) \\

  Environmental filtering & Position change (``position'') & When an increase in observations' trait similarity happens through some strong abiotic constraints \citep{cornwell2006trait}. For example, some trait combinations become more and more unlikely due to some environmental constraints (i.e. some regions of the trait space become unsuitable). & Metrics capturing changes in position (and richness) \\\\

  Competitive exclusion & Evenness change (``evenness'') & When functionally similar observations compete more intensively with one another than with functionally dissimilar observations (``competitive exclusion principle''; \citealt{hardin1960competitive}). At the extreme, it implies that only dissimilar individuals will coexist (``limiting similarity principle''; \citealt{macarthur1967limiting}). & Metrics capturing changes in divergence (but not richness and regularity) \\\\

@@

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.5\textwidth]{Figures/simulation_protocol_explained.pdf}
\caption{\scriptsize{Illustration of the simulation protocol used in this paper.
A) we simulated a pure birth-tree with a trait evolving under a Brownian Motion process until reaching 200 observations resulting a trait space of 2, 4, or 8 dimensions.
The first plot on row A illustrates one such simulation in one dimensions through time (with time on the horizontal axis and trait value for one dimension on the vertical axis).
The second plot on the same row shows the same results but with 2 dimensions of the trait represented here.
We then applied different stressors to the resulting trait space.
We represent the observations removed by the stressors in blue and the ones kept in the trait space in orange.
The first column (histograms) illustrates the distribution of observations in one dimension of the 2D, 4D or 8D traits with the overlap between the removed (blue) and kept (orange) observations in brown (note we did not measure any metric on 1D spaces).
The second column illustrate the trait in 2 dimensions.
We did not illustrate the data beyond 2 dimensions for traits in 4 or 8 dimensions.
We used the following stressors:
B) \textbf{null mechanism}: randomly removing a proportion of the observations;
C) \textbf{approximating equalising mechanism}: removing observations on the edge of the distribution;
D) \textbf{approximating facilitation mechanism}: removing observations to reduce the distance between pairs of observations (increasing local density; decreasing evenness);
E) \textbf{approximating filtering mechanism}: removing observations on one extreme of the distribution;
F) \textbf{approximating competition mechanism}: proportionally removing observations on the centre of the distribution.
The expected changes indicate the main type of metric affect by the stressor (following the aspects in \citealt{mammola2021concepts}).
G) we presented the results of the simulations as the scaled metric scores for the approximated mechanisms (C to F) relative to the null mechanism. The dot represents the median metric score and the thick and dashed lines respectively the 50\% and 95\% confidence intervals.}}
\label{Fig:simulations}
\end{figure}
\bigskip
